Tulai

### 2023-09-08 paper_tulai/paper_tulai.qmd
tidy up
WABI-Removed zero values from dataset

      TRUE ~ NA_character_
  ))
# Remove specific columns
tl1 %>% 
  select(-c("AREA", 
            "DEPTH (cm)", 
            "Raw Material", 
            "Colour/Grain/Opacity", 
            "Pattern"))

@@

sum_depth_sheet <- "https://docs.google.com/spreadsheets/d/1Q0QZESk412ZQLE24yPs6Rg-7Y9OMX63DHoXeIPNAIFM/edit#gid=0"

sum_depth_sheet <- read_sheet(sum_depth_sheet)

sum_depth_sheet <- sum_depth_sheet %>%
  mutate(area_layer_level = paste0(toupper(area), "_", layer, "_", level)) %>%
  relocate(area_layer_level, .after = level)

sum_depth_sheet <- sum_depth_sheet %>%
  mutate(retouch_freq = round(retouch_tool / total_lithics, 2),
         lithic_dens = round(total_lithics / volume, 2)
  )

write_sheet(sum_depth_sheet, ss = sheet_url)

@@

```{r}
# Create the ggplot
WABI_Level <- ggplot(sum_depth_sheet, 
            aes(x = lithic_dens,
                y = retouch_freq)) +

@@

```{r}
# Create the ggplot with y-axis on a log scale
WABI_Level <- ggplot(sum_depth_sheet, 
            aes(x = lithic_dens,
                y = retouch_freq)) +

  # Add points with different shapes based on area
  geom_point(aes(shape = area), size = 2) +

  # Add smoothing line
  geom_smooth(method = "lm", 
              se = TRUE, 
              color = "blue", 
              size = 0.5) +

  # Add regression equation, R-squared, and p-value
  stat_poly_eq(aes(label = paste(stat(eq.label), 
                                 stat(rr.label), 
                                 stat(p.value.label), 
                                 sep = "~~~")),
               formula = y ~ x, 
               parse = TRUE, 
               size = 2) +

  # Apply log scale to y-axis
  scale_y_log10() +

  # Add shape legend
  scale_shape_manual(values = c(16, 17, 25)) +

  # Customize theme
  theme(
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 12)
  )

# Print the plot
print(WABI_Level)

```{r}
# log-scale

WABI_Level <- ggplot(sum_depth_sheet, 
            aes(x = lithic_dens,
                y = retouch_freq)) +

  # Add smoothing line
  geom_smooth(method = "lm", 
              se = TRUE, 
              color = "blue", 
              size = 0.5) +

  # Add text labels for each point
  geom_text(aes(label = area_layer_level), 
            vjust = 1.5, 
            hjust = 0.5, 
            size = 2,
            check_overlap = TRUE) +

  labs(
    title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
    x = "Lithic Volumetric Density",
    y = "Retouch Frequency"
  ) +

  scale_y_log10() +

@@

# Create the ggplot
WABI_Level <- ggplot(sum_depth_sheet, 
            aes(x = lithic_dens,
                y = retouch_freq)) +

  # line with log scale

stat_smooth(method = "lm", 
            formula = y ~ log(x), 
            se = TRUE, aes(group = 1), 
            color = "blue", 
            size = 0.5)

  # Add text labels for each point
  geom_text(aes(label = area_layer_level), 
            vjust = -1, 
            hjust = -0.5, 
            size = 2.5) +

@@

```{r}
# Load required packages
library(ggplot2)
library(cowplot)

# Filter the data to exclude the value 0 and keep only blade, bladelet, and flake
tl4_filtered <- tl4_tool_Blank %>%
  filter(Blank2 %in% c("blade", "bladelet", "flake"))

# Plot Utilization by Blank2
plot_utilization <- ggplot(tl4_filtered, aes(Blank2, Utilization)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Utilization by Blank",
       x = "Blank2",
       y = "Utilization") +
  theme_minimal()

# Plot Retouch by Blank
plot_retouch <- ggplot(tl4_filtered, aes(Blank2, Retouch)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Retouch by Blank",
       x = "Blank2",
       y = "Retouch") +
  theme_minimal()

# Plot shine by Blank
plot_shine <- ggplot(tl4_filtered, aes(x = Blank2, y = shine2)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "shine by Blank",
       x = "Blank2",
       y = "shine2") +
  theme_minimal()

# Combine the plots using cowplot or patchwork
plot_combined <- cowplot::plot_grid(plot_utilization, plot_retouch, plot_shine, ncol = 3)

print(plot_combined)

#|label: fig_
#|fig-cap:  vs Debitage
#|fig-height: 5

# Filter data
tl4__blank %>% 
  filter(Blank2 %in% c('tp1', 'ta', 'd1')) %>% 
  select(Blank2, 
         `Width (mm)`, 
         `Thickness (mm)`, 
         `Length (mm)`) %>%
mutate(`Thickness (mm)` = as.numeric(`Thickness (mm)`))

# Pivot data to longer format
long_data <- tl3_dimention_blank %>% 
  pivot_longer(cols = c(`Width (mm)`, `Thickness (mm)`, `Length (mm)`), 
               names_to = "Dimension", 
               values_to = "Value")

# Create boxplot
ggplot(long_data, aes(x = area, y = Value)) +
  geom_boxplot() +
  geom_quasirandom(alpha = 0.01) +
  facet_wrap(~Dimension, scales = "free_y") +
  scale_y_log10() +
  theme_minimal()

```{r}
library(ggbeeswarm)
tl3_dimention_blank <- tl3 %>% 
  filter(area %in% tl3_area_tally_greater_than_n$area) %>% 
  pivot_longer(-area) %>% 
  ggplot() +
  aes(area,
      value) +
  geom_boxplot() +
  geom_quasirandom(alpha = 0.01) +
  facet_wrap( ~ name,
              scales = "free_y") +
  scale_y_log10() +
  theme_minimal()

```{r}
library(ggbeeswarm)
tl3 %>% 
  filter(area %in% tl3_area_tally_greater_than_n$area) %>% 
  pivot_longer(-area) %>% 
  ggplot() +
  aes(area,
      value) +
  geom_boxplot() +
  geom_quasirandom(alpha = 0.01) +
  facet_wrap( ~ name,
              scales = "free_y") +
  scale_y_log10() +
  theme_minimal()

```{r}
#| label: fig-Width-of-Artifacts-Area-A-by-depth
#| fig-cap: "Width of Artifacts by depth (Area A)"
#| fig-height: 5
tl %>%
  filter(AREA == "A") %>%
  select(
         `Width (mm)`,
         `DEPTH (cm)`) %>%
  ggplot() +
  aes(x = `DEPTH (cm)`,
      y = `Width (mm)`) +
  geom_boxplot() +
  theme_minimal()
```

```{r}
#| label: Width of Artifacts (Area A)
#| fig-cap: "Width of Artifacts (Area A)"
#| fig-height: 5

tl1 %>%
  filter(AREA == "TP1") %>%
  select(AREA, `Width (mm)`) %>%
  ggplot() +
  aes(x = `Width (mm)`) +
  geom_histogram(binwidth = 2) +
  labs(x = "Width of Debitage (mm)", 
       y = "Frequency", 
       title = "Histogram of Width for Area TP1" )+
  theme_minimal()
```

WABI:

```{r}
# regression plot-layer
library(ggplot2)
library(ggpmisc)

@@

@@

@@

# Creating a biplot
fviz_pca_biplot(res_pca)
```

@@

# Get the loadings (coefficients) of original variables for each principal component
loadings_matrix <- res_pca$var$coord

@@

@@

```{r}
#|fig-cap: Contribution of variables for each of the tow relevant PCA dimensions

@@

```{r}
# Create the initial biplot without individual labels
p <- fviz_pca_biplot(res_pca_area, label = "none", 
                     repel = TRUE, labelsize = 4)

# Add the text labels for the areas using their coordinates from the PCA result
p + geom_text(aes(x = res_pca_area$ind$coord[, 1], 
                  y = res_pca_area$ind$coord[, 2], 
                  label = pca_final_table_area$area), 
             size = 3, nudge_x = 0.1, nudge_y = 0.1)

@@

OK

```{r}
#| label: fig-1-layer
#| fig-cap: "Type of Removals"
#| fig-height: 5

#Type of blank based on Layers

library(ggbeeswarm)

tl4 %>% 
  filter(!is.na(layer) & Blank2 %in% c('bladelet', 
                                       'blade', 
                                       'flake')) %>%

  # Create bar plot
  ggplot() +
  aes(x = layer, 
      fill = Blank2) +
  geom_bar(position = "dodge") +
  labs(x = "Area", 
       y = "Number", 
       fill = "Blank Type") +
  theme_minimal()

```{r}
#Type of core based on Layers

library(ggbeeswarm)

tl4 %>% 
  filter(!is.na(layer) & 
           `core-typology` %in% c('bullet','prismatic','pyramid')) %>%

  # Create bar plot
  ggplot() +
  aes(x = layer, 
      fill = `core-typology`) +
  geom_bar(position = "dodge") +
  labs(x = "Area", 
       y = "Number", 
       fill = "Blank Type") +
  theme_minimal()
```

consistency in dimention:

```{r}
#|label: fig_2
#|fig-cap: "Boxplot for each Dimention"
#|fig-height: 5

library(ggplot2)
library(dplyr)
library(tidyr)

# Set the value for n
n <- 100

# Tally by area
tl_final_area_tally <- tl_final_area %>%
  group_by(area) %>%
  tally()

# Filter areas with more than n rows
tl_final_area_tally_greater_than_n <- tl_final_area_tally %>%
  filter(n >= n)

# Filter data based on areas with more than n rows
tl_final_area_filtered <- tl_final_area %>%
  filter(area %in% tl_final_area_tally_greater_than_n$area) %>%
  filter(!is.na(blank) & blank != "")

# Select and mutate dimensions
tl_final_area_selected <- tl_final_area_filtered %>%
  select(area, width, thickness, length) %>%
  mutate(
    width = as.numeric(width),
    thickness = as.numeric(thickness),
    length = as.numeric(length)
  )

# Pivot data to longer format
tl_final_area_long <- tl_final_area_selected %>%
  pivot_longer(cols = c(width, thickness, length), 
               names_to = "Dimension", 
               values_to = "Value")

# Create the box plot
ggplot(tl_final_area_long, aes(x = area, y = Value)) +
  geom_boxplot() +
  facet_wrap(~Dimension, scales = "free") +
  labs(x = "Area", y = "Value") +
  theme_minimal()

```{r}
#|label: fig_3
#|fig-cap: Dimention By Area
#|ffig-height: 5

library(ggplot2)
library(dplyr)
library(tidyr)
library(ggbeeswarm)

# Create boxplot
ggplot(tl_final_area_long, 
       aes(x = area, y = Value)) +
  geom_boxplot() +
  geom_quasirandom(alpha = 0.01) +
  facet_wrap(~Dimension, scales = "free_y") +
  scale_y_log10() +
  theme_minimal()

```{r}
#|label: fig_2-layer
#|fig-cap: "Boxplot for each Dimention"
#|fig-height: 5

n <- 100

# Tally count of each layer
tl_final_layer_tally <- tl_final_area %>% 
  group_by(layer) %>% 
  tally()

# Filter layers with more than n rows
tl_final_layer_tally_greater_than_n <- tl_final_layer_tally %>% 
  filter(n >= n)

# Filter data for layers with more than n rows
tl_final_layer_filtered <- tl_final_area %>% 
  filter(layer %in% tl_final_layer_tally_greater_than_n$layer)

# Select and mutate
tl_final_layer_selected <- tl_final_layer_filtered %>%
  select(layer, 
         `width`, 
         `thickness`, 
         `length`) %>%
  mutate(across(c
                (
           `width`, 
           `thickness`, 
           `length`), 
           as.numeric)
         )

# Pivot longer
tl_final_layer_long <- tl_final_layer_selected %>%
  pivot_longer(cols = 
                 c(width, 
                   thickness, 
                   length), 
               names_to = 
                 "Dimension", 
               values_to = 
                 "Value") %>%
  drop_na(Value)

# Create box plot
ggplot(tl_final_layer_long, 
       aes(layer, 
           Value)) +

  geom_boxplot() +
  facet_wrap(~Dimension, 
             scales = "free") +
  labs(x = "Layer", 
       y = "Value") +

   theme_minimal()

```{r}
#|label: fig_3-layer
#|fig-cap: Dimension By Layer
#|fig-height: 5

# Create boxplot
ggplot(tl_final_layer_long, 
       aes(x = layer, y = Value)) +
  geom_boxplot() +
  geom_quasirandom(alpha = 0.01) +
  facet_wrap(~Dimension, scales = "free_y") +
  scale_y_log10() +
  theme_minimal()

```{r}
#|label: fig_4-width
#|fig-cap: "Width of Artifacts (Area A)"
#| fig-height: 5
tl_final_area %>%
 filter(area == "ta") %>%
  select(area, 
         width) %>%
  ggplot() +
  aes(x = width) +
  geom_histogram(binwidth = 2) +
  labs(x = "Width of Debitage (mm)", 
       y = "Frequency" )+
  theme_minimal()
```

```{r}
#|label: fig_5-width
#|fig-cap: "Width of Artifacts (Area A, D1, TP1)"
#|fig-height: 5

tl_final_area %>%
  filter(area %in% 
           c("ta", "d1", "tp1")) %>%
  mutate(width = as.numeric(width)) %>%  # Explicitly convert to numeric

   select(area, 
          `width`) %>%
  ggplot() +

  aes(x = width) +  # Make sure the aesthetic mapping is correct

  geom_histogram(binwidth = 2) +

  labs(x = "Width of Debitage (mm)", 
       y = "Frequency" ) +

  facet_wrap(~area, 
             ncol = 1) +

  theme_minimal()

```{r}
#|label: fig_5-width-layer
#|fig-cap: "Width of Artifacts (Area A, D1, TP1)"
#|fig-height: 5

tl_final_area %>%
  filter(layer %in% c("a", 
                      "c", 
                      "e or d", 
                      "e", 
                      "d")) %>%
  mutate(width = as.numeric(width)) %>%  # Convert to numeric
  select(layer, width) %>%
  ggplot() +
  aes(x = width) +
  geom_histogram(binwidth = 2) +
  labs(x = "Width of Debitage (mm)", y = "Frequency") +
  facet_wrap(~layer, ncol = 1) +
  theme_minimal()

```{r}
#|label: fig_6-thickness-layer
#|fig-cap: "Thickness of Artifacts (Layers A, C, E or D)"
#|fig-height: 5

tl_final_area %>%
 filter(layer %in% c("a", 
                     "c", 
                     "e or d", 
                     "d", 
                     "e")) %>%
 mutate(thickness = as.numeric(thickness)) %>% 
  select(layer, 
         thickness) %>%
  ggplot() +
  aes(x = thickness) +
  geom_histogram(binwidth = 2) +
  labs(x = "Thickness of Debitage (mm)", 
       y = "Frequency" )+
  facet_wrap(~layer , ncol = 1) +
  theme_minimal()

```{r}
#|label: fig_7
#|fig-cap: "Length of Artifacts (Area A, D1, TP1)"
#|fig-height: 5

tl3 %>%
  filter(Blank_part == "complete", area %in% c("ta", "tp1", "d1")) %>%
  mutate(`Length (mm)` = as.numeric(`Length (mm)`)) %>%
  select(area, 
         `Length (mm)`) %>%
  ggplot() +
  aes(x = `Length (mm)`) +
  geom_histogram(binwidth = 2) +
  labs(x = "Length of Debitage (mm)", 
       y = "Frequency") +
  facet_wrap(~area, ncol = 1) +
  theme_minimal()

```{r}
#| label: fig-7-length-layer
#| fig-cap: Length of Artefact based on layers
#| fig-height: 5
tl3 %>% 
  filter(layer %in% c("layer a" , "layer d" , "layer e" , "layer c" , "layer e or d")) %>% 
  mutate(`Length (mm)` = as.numeric(`Length (mm)`)) %>% 
  select(layer, `Length (mm)`) %>% 
  ggplot(aes(x = `Length (mm)`)) + 
  geom_histogram(binwidth = 2) +
  labs(x = "Length of Debitage (mm)" ,
       y = "Frequency") + 
  facet_wrap(~layer, ncol = 1) + 
  theme_minimal()

```{r}
#|label: NA 
#|fig-cap: "tool typology"
#|fig-height: 5

```{r}
library(ggplot2)

# Calculate the counts for debitage and 
counts <- tl4_tool_Blank %>%
  group_by(debitage, ) %>%
  summarise(count = n())

# Create the bar plot
ggplot(counts, 
       aes(x = factor(debitage), 
           y = count, 
           fill = factor(debitage))) +

  geom_bar(stat = "identity", 
           position = "dodge") +

  # Move the labels slightly above the bars
  geom_text(aes(label = count), 
            vjust = -0.5) +

  geom_hline(data = counts %>% 
               filter( == 1), 
             aes(yintercept = count),
             linetype = "dashed", 
             color = "red") +
  labs(title = "Debitage vs. ",
       x = "",
       y = "Numbers") +
  theme_minimal()

```{r}
# Calculate the counts for debitage and 
counts <- tl4_tool_Blank %>%
  group_by(Retouch = 1,  =1) %>%
  summarise(count = n())

# Create the bar plot
ggplot(counts, 
       aes(x = factor( = 1), 
           y = count, 
           fill = factor( = 1))) +

  geom_bar(stat = "identity", 
           position = "dodge") +

  # Move the labels slightly above the bars
  geom_text(aes(label = count), 
            vjust = -0.5) +

  geom_hline(data = counts %>% 
               filter( == 1), 
             aes(yintercept = count),
             linetype = "dashed", 
             color = "red") +
  labs(title = " vs. formal ",
       x = "",
       y = "Numbers") +
  theme_minimal()
```

```{r}
#|label: fig_
#|fig-cap: ""
#| fig-height: 5
tl4_tool_Blank %>%
  filter(Blank2 %in% c("bladelet", "blade", "flake")) %>%
  select(Blank2, shine2 = 1) %>%
  ggplot() +
  aes(x = shine2) +
  geom_bar() +
  labs(x = "shine", 
       y = "Frequency" )+
  theme_minimal()

tl4_tool_Blank %>%
  filter(Blank2 %in% c("bladelet", "blade", "flake")) %>%
  select(Blank2, Retouch = 1) %>%
  ggplot() +
  aes(x = Retouch) +
  geom_bar() +
  labs(x = "Retouch", 
       y = "Frequency" )+
  theme_minimal()

tl4_tool_Blank %>%
  filter(Blank2 %in% c("bladelet", "blade", "flake")) %>%
  select(Blank2, Utilization = 1) %>%
  ggplot() +
  aes(x = Utilization) +
  geom_bar() +
  labs(x = "Utilization", 
       y = "Frequency" )+
  theme_minimal()

```{r}
library(ggplot2)
library(patchwork)

plot_shine <- tl4_tool_Blank %>%
  filter(Blank2 %in% c("bladelet", 
                       "blade", 
                       "flake")) %>%
  select(Blank2, shine2 = 1) %>%
  ggplot() +
  aes(x = shine2) +
  geom_bar() +
  labs(x = "shine", y = "Frequency") +
  theme_minimal()

plot_retouch <- tl4_tool_Blank %>%
  filter(Blank2 %in% c("bladelet", 
                       "blade", 
                       "flake")) %>%
  select(Blank2, 
         Retouch = 1) %>%
  ggplot() +
  aes(x = Retouch) +
  geom_bar() +
  labs(x = "Retouch", 
       y = "Frequency") +
  theme_minimal()

plot_utilization <- tl4_tool_Blank %>%
  filter(Blank2 %in% c("bladelet", "blade", "flake")) %>%
  select(Blank2, Utilization = 1) %>%
  ggplot() +
  aes(x = Utilization) +
  geom_bar() +
  labs(x = "Utilization", y = "Frequency") +
  theme_minimal()

# Combine the plots using patchwork
plot_combined <- plot_shine + plot_retouch + plot_utilization
plot_combined

>>>>

      TRUE ~ NA_character_
  ))

# Remove specific columns
tl1 %>% 
  select(-c("AREA", 
            "DEPTH (cm)", 
            "Raw Material", 
            "Colour/Grain/Opacity", 
            "Pattern"))

@@

google_sheet_url <- "https://docs.google.com/spreadsheets/d/1Q0QZESk412ZQLE24yPs6Rg-7Y9OMX63DHoXeIPNAIFM/edit#gid=0"

sum_depth_sheet <- read_sheet(google_sheet_url)

sum_depth_sheet <- sum_depth_sheet %>%
  mutate(area_layer_level = paste0(toupper(area), "_", layer, "_", level)) %>%
  relocate(area_layer_level, .after = last_col()) %>%
  mutate(
    retouch_freq = round(retouch_tool / total_lithics, 2),
    lithic_dens = round(total_lithics / volume, 2)
  )

write_sheet(sum_depth_sheet, ss = google_sheet_url)

@@

```{r}
# WABI_LEVEL
filtered_data <- sum_depth_sheet %>% filter(lithic_dens > 0 & retouch_freq > 0)

WABI_Level <- ggplot(filtered_data, 
            aes(x = lithic_dens,
                y = retouch_freq)) +

@@

```{r}
#  WABI-level-log scale

filtered_data <- sum_depth_sheet %>% filter(lithic_dens > 0 & retouch_freq > 0)

WABI_Level <- ggplot(filtered_data,  
            aes(x = lithic_dens,
                y = retouch_freq)) +

  geom_point(color = "black", size = 1.5) +

  # Add text labels 
  geom_text(aes(label = area_layer_level), 
            vjust = 1.5, 
            hjust = 0.5, 
            size = 2,
            check_overlap = TRUE) +

  #regression equation, R-squared, and p-value
  stat_poly_eq(aes(label = paste(stat(eq.label), 
                                  stat(rr.label), 
                                  stat(p.value.label), 
                                  sep = "~~~")),
               formula = y ~ x, 
               parse = TRUE, 
               size = 2) +

  # titles and labels
  labs(
    title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
    x = "Lithic Volumetric Density",
    y = "Retouch Frequency"
  ) +

  # Apply log scales
  scale_x_log10() +
  scale_y_log10() +

@@

@@

#WABI:

# regression plot-layer
library(ggplot2)
library(ggpmisc)

@@

print(final_table)

@@

# contributions of variables to the second principal component
fviz_contrib(res_pca,
             axes = 2,   
             choice = "var")

@@

#  biplot
fviz_pca_biplot(res_pca)
```

@@

```{r}
# Run PCA
library(FactoMineR)
library(factoextra)

# Get the coefficients of original variables for each principal component
loadings_matrix <- res_pca$var$coord

print(loadings_matrix)

@@

print(df)

@@

@@

```{r}
# biplot 
p <- fviz_pca_biplot(res_pca_area, label = "none", 
                     repel = TRUE, labelsize = 4)

p + geom_text(aes(x = res_pca_area$ind$coord[, 1], 
                  y = res_pca_area$ind$coord[, 2], 
                  label = pca_final_table_area$area), 
             size = 3, nudge_x = 0.1, nudge_y = 0.1)

@@
